---
title: Blue Nile + Data Analysis = Diamond Deals
date: 2015-11-11
---

# Introduction

# Downloading the data

# Modeling price as a function of the Four C's

# Identifying diamond deals

This post is aimed at everyone who wants to buy a diamond and is excited by the prospect of data mining for good deals on diamonds. If you're still reading, you're in a niche audience!

Are you shopping for a diamond? This is a guide on how to download data from 

> Founded in 1999, Blue Nile has grown to become the largest online
> retailer of certified diamonds and fine jewelry. Internet Retailer
> Magazine reports Blue Nile is bigger than the next three largest
> online jewelers combined.


When browsing diamonds on Blue Nile, I noticed that requests are made to a Solr
database containing information about each diamond on Blue
Nile. This is awesome. On November 6, 2015, I downloaded the data on
all 98,886 round diamonds on Blue Nile! My understanding of diamond
pricing is based on Blue Nile's excellent
[diamond buying guide][buying-guide]. The four C's (cut, color,
clarity, and carat weight) are the most important characteristics when
buying a diamond. Let's model price as a function of those four
characteristics and use that model to find potential deals.

```{r, message=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
library(msm)
library(knitr)
library(grDevices)

opts_chunk$set(echo=FALSE)
options(stringsAsFactors=TRUE)
diamonds <- read.csv('rounds.csv')
diamonds$tenths <- floor((10 * diamonds$carat) %% 10)

# Set up dummy variables
values <- list(
  cut=c('Good', 'Very Good', 'Ideal', 'Signature Ideal'),
  color=c('J', 'I', 'H', 'G', 'F', 'E', 'D'),
  clarity=c('SI2', 'SI1', 'VS2', 'VS1', 'VVS2', 'VVS1', 'IF', 'FL')
)
for (x in names(values)) {
  vals <- values[[x]]
  for (i in 2:length(vals)) {
    diamonds[[paste0(x, i)]] <- as.numeric(diamonds[[x]] == vals[i])
    diamonds[[x]] <- ordered(diamonds[[x]], levels=vals)
  }
}
dummies <- grep('\\d', names(diamonds), value=TRUE)

stats <- lapply(diamonds[, c('price', 'carat', dummies)], function(x)
  c(min=min(x), median=median(x), mean=mean(x), max=max(x))
)
```

```{r big_picture}
# Use some uniform noise to smooth out distribution of diamond weights.
# Diamond weight is rounded to 1/100 of a carat.
diamonds$noise <- runif(nrow(diamonds), 0, 0.01)

carats <- 2^(-2:4)
prices <- 1000 * 5^(0:4)
price_labels <- paste0('$', prices / 1000, ',000')
(
    ggplot(diamonds %>% filter(carat <= 8), aes(x=log(carat + noise), y=log(price))) +
    stat_smooth(method='lm') +
    geom_point(size=0.3) +
    scale_x_continuous(breaks=log(carats), labels=carats) +
    scale_y_continuous(breaks=log(prices), labels=price_labels) +
    ggtitle('Diamond price as a function of diamond weight') +
    ylab("Price") +
    xlab('Carat Weight') +
    theme_bw() +
    theme(panel.grid.minor.x=element_blank(), panel.grid.minor.y=element_blank())
)
```

```{r zoomed_in}
carats <- seq(1, 1.9, 0.1)
prices <- 1000 * 2^(2:5)
price_labels <- paste0('$', prices / 1000, ',000')
diamonds <- diamonds %>% filter(1 <= carat & carat < 2)
(
    ggplot(diamonds, aes(x=log(carat + noise), y=log(price), color=color)) +
    geom_point(size=0.4) +
    stat_smooth(method='lm') +
    scale_x_continuous(breaks=log(carats), labels=carats) +
    scale_y_continuous(breaks=log(prices), labels=price_labels) +
    ggtitle('Diamond price as a function of weight and color') +
    ylab("Price") +
    xlab('Carat Weight') +
    theme_bw() +
    theme(panel.grid.minor.x=element_blank(), panel.grid.minor.y=element_blank())
)

my_formula <- paste('log(price) ~ log(carat) +',
  paste(dummies, collapse='+')
)
fit <- lm(my_formula, data=diamonds)
summary(fit)
```

We just fit the following model
$$
\log(price_i) = \alpha + \beta^0 \log(carat_i) + \beta^1_{cut(i)} + \beta^2_{color(i)} + \beta^3_{clarity(i)} + \epsilon_i
$$
Exponentiating both sides of the equation we have a multiplicative relationship between price and the Four C's:
$$
price_i = \exp(\alpha) \cdot carat_i^{\beta^0} \cdot \exp\left(\beta^1_{cut(i)}\right) \cdot \exp\left(\beta^2_{color(i)}\right) \cdot \exp\left(\beta^3_{clarity(i)}\right) \cdot \exp(\epsilon_i)
$$
Exponentiating the coefficients from our regression model we can get estimates of the price multipliers associated with different diamond characteristics. These multipliers can help a shopper decide what type of diamond to look at. Is a G-color diamond worth `r round(exp(coef(fit)['color4']), 2)` times the price of a J-color diamond?

```{r coefplot}
l <- lapply(dummies, function(s) {
  beta <- coef(fit)[[s]]
  variance <- vcov(fit)[s, s]
  new_variance <- deltamethod(~ exp(x1), beta, variance)
  c(multiplier=exp(beta), variance=new_variance)
})
multipliers <- data.frame(do.call(rbind, l))
multipliers$variable <- unlist(sapply(names(values), function(x) paste(values[[x]][-1], '-', x)))
multipliers$y <- nrow(multipliers) - 1:nrow(multipliers)
diff <- qnorm(0.95, mean=0, sd=sqrt(multipliers$variance))
multipliers$low <- multipliers$multiplier - diff
multipliers$high <- multipliers$multiplier + diff
(
    ggplot(multipliers, aes(x=multiplier, y=y)) +
    geom_point() +
    scale_y_continuous(breaks=multipliers$y, labels=multipliers$variable) +
    theme_bw() +
    geom_errorbarh(aes(xmax=high, xmin=low), height=0) +
    ylab('') +
    xlab('Price Multiplier') +
    ggtitle('Price multipliers for cut, color, and clarity types')
)
```

```{r}
yhat <- exp(predict(fit))
diamonds$markup <- (diamonds$price / yhat - 1) * 100
diamonds %>% arrange(markup) %>% select(markup, carat, cut, color, clarity, price) %>% head(20)
```

[blue-nile]: http://www.bluenile.com/
[solr]: http://lucene.apache.org/solr/
[buying-guide]: http://www.bluenile.com/education/diamonds/
[delta-method]: http://www.ats.ucla.edu/stat/r/faq/deltamethod.htm
